Visualising CHD Pathway Inequalities by ethnicity - DRAFT

### 2024-04-29 draft_report.qmd
Additional references and tweaked chart heights

Ethnicity, unlike deprivation, is not an ordered variable. The approach in our previous work used the relative index of inequality to measure the degree of inequality. This measure relies on the ordered quality of the socio-economic deprivation variable. To handle the categorical nature of the ethnicity variable we have used the relative index of disparity to indicate the extent to which the rate of an activity or event varies across groups. The index estimates the proportion of events (e.g., admissions) that would need to be redistributed between clusters in order that event rates follow levels of need. Further detailed explanations of the methods used in the analysis can be found in the appendix.

@@

::: {.callout-note appearance="minimal"}
Work led by Professor Sir Michael Marmot and colleagues in the late 1980s revealed that first-generation South Asians living in the UK have a higher rate of coronary heart disease and diabetes compared to White Europeans[^4]. This gives rise to the question, why are the death rates lower in the more diverse clusters? There are a number of factors to consider in understanding this including; the age/sex profile of the clusters, the methodology for calculating the CHD synthetic prevalence (the measure of need in this project), and age and sex standardised mortality rates (ASMR) compared to the project CHD mortality rates. These are examined further in the appendix.
:::

[^4]: <https://www.bhf.org.uk/what-we-do/our-research/research-successes/ethnicity-and-heart-disease>

@@

```{r}
tar_read(ci_iod_chart)
```

@@

A previous Strategy Unit report "Strategies to reduce inequalities in access to planned hospital procedures"[^5] highlights in chapter 1, three of the many routes from inequity to equity: levelling-up, levelling-down and zero-sum redistribution. Using the zero-sum redistribution route as an example, along with the metric flu vaccinations for patients aged 65+, the illustration below shows how equity could be achieved by increasing and decreasing the number of vaccinations delivered in each cluster to give the same activity rate per cluster.

[^5]: <https://www.midlandsdecisionsupport.nhs.uk/knowledge-library/strategies-to-reduce-inequalities-in-access-to-planned-hospital-procedures/>

@@

|  Cluster  | Number of vaccinations | CHD synthetic prevalence | Rate (No. of vaccinations / CHD synthetic prevalence) | Global rate | Change in number of vaccinations to match global rate |
|:---------:|-----------------------:|-------------------------:|:-----------------------------------------------------:|:-----------:|:-----------------------------------------------------:|
|           |                    *n* |                      *d* |                        *r=n/d*                        | *gr=𝚺n/𝚺d*  |                     *(gr-r) x d*                      |
|     1     |              3,356,002 |                1,228,453 |                         2.73                          |    1.92     |                      -1,001,284                       |
|     2     |              2,638,698 |                1,124,040 |                         2.35                          |    1.92     |                       -484,121                        |
|     3     |              1,862,430 |                1,181,796 |                         1.58                          |    1.92     |                        402,855                        |
|     4     |                476,616 |                  571,049 |                         0.83                          |    1.92     |                        617,979                        |
|     5     |                332,728 |                  415,950 |                         0.80                          |    1.92     |                        464,571                        |
| **Total** |          **8,666,474** |            **4,521,288** |                         1.92                          |             |                         **0**                         |

@@

```{r}
regional_charts <- tar_read(regional_charts)
regional_charts[1]
```

@@

```{r}
regional_charts[2]
```

@@

```{r}
regional_charts[3]
```

@@

```{r}
regional_charts[4]
```

@@

```{r}
regional_charts[5]
```

@@

```{r}
regional_charts[6]
```

@@

```{r}
regional_charts[7]
```
:::

@@

The methodology used for this metric follows, as far as possible, that detailed by NHS Digital[^6]. The readmissions counted relate to those who had an elective PCI / CABG in the period 1/3/2022 to 28/2/2023.

[^6]: <https://digital.nhs.uk/data-and-information/publications/statistical/ccg-outcomes-indicator-set/specifications/3.2-emergency-readmissions-within-30-days-of-discharge-from-hospital_1_4>

@@

Public Health England (PHE) CHD synthetic prevalence estimates data was previously available from OHID via Fingertips. This data has now been removed from Fingertips and archived, however OHID supplied it via email for this project following a request. This data is for 7564 practices from 2015, of which 6297 are current practices. In order to assign a prevalence for practices with missing data a nearest neighbours methodology was used to impute the prevalence by taking an average of the prevalence of each practice within a 1.5km radius of the practice without data. Further details, including code with sample data, for this methodology are available in a blog post on the Strategy Unit Data Science website[^7].

[^7]: <https://the-strategy-unit.github.io/data_science/blogs/posts/2024-01-17_nearest_neighbour.html>

@@

Since metric data is mainly from the 2022/23 year, the October 2022 list size data was taken from the NHS Digital Patients Registered at a GP practice dataset[^8] as this is the data closest to the mid point in the 22/23 year. The GP to ICB mapping file was also downloaded and joined to the list size to easily create analysis by ICB and region.

[^8]: <https://digital.nhs.uk/data-and-information/publications/statistical/patients-registered-at-a-gp-practice/october-2022>

@@

The 2021 Census Ethnicity data was downloaded from the .gov.uk website.[^9] Summary details of the census ethnicity data is also available and this includes descriptions of the standardised list of ethnic groups and the differences from the 2011 ethnic groups.[^10]

[^9]: <https://www.ethnicity-facts-figures.service.gov.uk/uk-population-by-ethnicity/demographics/age-groups/latest/#download-the-data>

[^10]: <https://www.ethnicity-facts-figures.service.gov.uk/uk-population-by-ethnicity/demographics/age-groups/latest/>

@@

Further details about item editing and imputation processes for Census 2021, England and Wales are available on the ONS website. This details techniques used by ONS with the aim of arriving at a fully populated clean database. These techniques include manual imputation and nearest neighbour donor imputation. For the England and Wales Census 2021 the imputation rate for ethnic group was 1.3%.[^11]

[^11]: <https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/methodologies/itemeditingandimputationprocessforcensus2021englandandwales>

@@

The K-medoids clustering was performed using the {cluster} package[^12] in R using partitioning (clustering) of the data into k clusters "around medoids" (PAM), which is a more robust version of K-means. The algorithm is based on the search for k representative objects or medoids among the observations of the dataset. After finding a set of k medoids, k clusters are constructed by assigning each observation to the nearest medoid. The goal is to find k representative objects which minimise the sum of the dissimilarities of the observations to their closest representative object.[^13]

[^12]: Maechler M, Rousseeuw P, Struyf A, Hubert M, Hornik K (2023). *cluster: Cluster Analysis Basics and Extensions*. R package version 2.1.6 — For new features, see the 'NEWS' and the 'Changelog' file in the package source), [https://CRAN.R-project.org/package=cluster](https://cran.r-project.org/package=cluster).

[^13]: <https://cran.r-project.org/web/packages/cluster/cluster.pdf>

@@

A report by Pearcy and Keppel (2002) presents the Index of Disparity as a summary measure of disparity across population groups and advocates its use for groups defined in terms of categories such as ethnicity.[^14]

[^14]: <https://journals.sagepub.com/doi/epdf/10.1093/phr/117.3.273>

@@

> "Practitioners need to be aware that statistical significance differs from practical importance in that statistical significance is highly dependent upon sample size. For a large sample, a statistically significant result is likely to be obtained even when the actual magnitude of an effect is small and of little or no practical importance. On the other hand, for a small sample, it is quite likely that insufficient evidence of a statistically significant result will be obtained – even when there is, indeed, an effect of practical importance." (Hahn, G. J., Doganaksoy, N., Meeker, W. Q., 2019)[^15]

[^15]: <https://academic.oup.com/jrssig/article/16/4/20/7038025>

@@

An explanation of bootstrap theory and confidence intervals is available in Practical Statistics for Data Scientists (Bruce, P. and Bruce, A., 2017)[^16]

[^16]: <https://www.oreilly.com/library/view/practical-statistics-for/9781491952955/>

@@

| Cluster       | CHD Prevalence (%) |  List size | CHD Prevalence | CHD Deaths (% of list size) | Activity rate (CHD Deaths/Prevalence) |
|---------------|:------------------:|-----------:|---------------:|:---------------------------:|:-------------------------------------:|
| Least diverse |        8.7%        | 14,000,000 |      1,218,000 |        6000 (0.043%)        |                0.00492                |
| Most diverse  |       10.6%        |  4,000,000 |        424,000 |        1714 (0.043%)        |                0.00404                |

@@

| Cluster | Population | ASMR - All cause | ASMR - CHD | Rate - All Cause (All deaths / Prevalence) | Rate - CHD (CHD deaths / Prevalence) |
|:-------:|-----------:|-----------------:|-----------:|-------------------------------------------:|-------------------------------------:|
|    1    | 16,914,882 |            913.5 |       93.9 |                                     1188.4 |                                123.1 |
|    2    | 15,788,339 |            894.3 |       87.6 |                                     1029.5 |                                100.5 |
|    3    | 16,416,864 |            935.4 |       93.3 |                                      813.7 |                                 80.6 |
|    4    |  7,883,280 |            773.7 |       82.2 |                                      431.4 |                                 45.9 |
|    5    |  4,961,545 |            868.2 |      105.8 |                                      511.1 |                                 62.7 |

@@

In July 2021 ONS produced an experimental analysis of ethnic differences in life-expectancy and cause-specific mortality rates[^17]. The analysis found that White and Mixed ethnic groups had lower life expectancy than all other ethnic groups, while the Black African ethnic group had statistically significant higher life expectancy than most groups. However, the analysis also found that ischaemic heart disease mortality was highest in the Bangladeshi, Pakistani and Indian ethnic groups and lowest among Black ethnic groups. The analysis revealed complex patterns in life expectancy and mortality by ethnic group. Potential reasons are offered for the higher life expectancy found in Black African and Asian Other ethnic groups including that they contain a greater proportion of more recent migrants than other ethnic groups and it also cites previous research showing that people who migrate tend to be healthier than others. The ONS report also suggests that Asian and Black ethnic groups engage less in harmful health-related behaviours. These findings show some similarities to research in the USA into what has been termed the Hispanic paradox, where most Hispanic groups are characterised by low economic status and a high prevalence of cardiovascular disease risk factors, but better than expected health and mortality outcomes[^18]. Possible explanations for the Hispanic paradox include the healthy migrant hypothesis (those who migrate tend to be younger and healthier), the presence of nuclear families and high levels of social support, and dietary factors.

[^17]: <https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/lifeexpectancies/articles/ethnicdifferencesinlifeexpectancyandmortalityfromselectedcausesinenglandandwales/2011to2014>

[^18]: <https://www.sciencedirect.com/science/article/abs/pii/S0033062014001339?via%3Dihub>

@@

All the code used in this project is available on the Strategy Unit GitHub[^19].

[^19]: <https://github.com/The-Strategy-Unit/bhf_chd-ethnicity>

>>>>

Ethnicity, unlike deprivation, is not an ordered variable. The approach in our previous work used the relative index of inequality to measure the degree of inequality. This measure relies on the ordered quality of the socio-economic deprivation variable. To handle the categorical nature of the ethnicity variable we have used the relative index of disparity[^4] to indicate the extent to which the rate of an activity or event varies across groups. The index estimates the proportion of events (e.g., admissions) that would need to be redistributed between clusters in order that event rates follow levels of need. Further detailed explanations of the methods used in the analysis can be found in the appendix.

[^4]: *Pearcy J, Keppel K, A Summary Measure of Health Disparity, Public Health Reports, Vol 117, May-June 2002*

    [*https://open.umich.edu/sites/default/files/downloads/PublicHealthRep-Pearcy.pdf*](https://gbr01.safelinks.protection.outlook.com/?url=https%3A%2F%2Fopen.umich.edu%2Fsites%2Fdefault%2Ffiles%2Fdownloads%2FPublicHealthRep-Pearcy.pdf&data=05%7C02%7Cjacqueline.grout1%40nhs.net%7Ca9c82e4d3cc94a72de8408dc68333fc0%7C37c354b285b047f5b22207b48d774ee3%7C0%7C0%7C638499816545595235%7CUnknown%7CTWFpbGZsb3d8eyJWIjoiMC4wLjAwMDAiLCJQIjoiV2luMzIiLCJBTiI6Ik1haWwiLCJXVCI6Mn0%3D%7C0%7C%7C%7C&sdata=ddk0rdD6HnQ1524JNCsTfr3AoiBePESeckGAAnQPSA4%3D&reserved=0)

@@

::: {.callout-note appearance="minimal"}
Work led by Professor Sir Michael Marmot and colleagues in the late 1980s revealed that first-generation South Asians living in the UK have a higher rate of coronary heart disease and diabetes compared to White Europeans[^5]. This gives rise to the question, why are the death rates lower in the more diverse clusters? There are a number of factors to consider in understanding this including; the age/sex profile of the clusters, the methodology for calculating the CHD synthetic prevalence (the measure of need in this project), and age and sex standardised mortality rates (ASMR) compared to the project CHD mortality rates. These are examined further in the appendix.
:::

[^5]: <https://www.bhf.org.uk/what-we-do/our-research/research-successes/ethnicity-and-heart-disease>

@@

```{r}
#| out-height: 600px

tar_read(ci_iod_chart)
```

@@

A previous Strategy Unit report "Strategies to reduce inequalities in access to planned hospital procedures"[^6] highlights in chapter 1, three of the many routes from inequity to equity: levelling-up, levelling-down and zero-sum redistribution. Using the zero-sum redistribution route as an example, along with the metric flu vaccinations for patients aged 65+, the illustration below shows how equity could be achieved by increasing and decreasing the number of vaccinations delivered in each cluster to give the same activity rate per cluster.

[^6]: <https://www.midlandsdecisionsupport.nhs.uk/knowledge-library/strategies-to-reduce-inequalities-in-access-to-planned-hospital-procedures/>

@@

|  Cluster  | Number of vaccinations | CHD synthetic prevalence | Rate (No. of vaccinations / CHD synthetic prevalence) | Global rate | Change in number of vaccinations to match global rate |
|:---------:|----------:|----------:|:------------:|:---------:|:------------:|
|           |                    *n* |                      *d* |                        *r=n/d*                        | *gr=𝚺n/𝚺d*  |                     *(gr-r) x d*                      |
|     1     |              3,356,002 |                1,228,453 |                         2.73                          |    1.92     |                      -1,001,284                       |
|     2     |              2,638,698 |                1,124,040 |                         2.35                          |    1.92     |                       -484,121                        |
|     3     |              1,862,430 |                1,181,796 |                         1.58                          |    1.92     |                        402,855                        |
|     4     |                476,616 |                  571,049 |                         0.83                          |    1.92     |                        617,979                        |
|     5     |                332,728 |                  415,950 |                         0.80                          |    1.92     |                        464,571                        |
| **Total** |          **8,666,474** |            **4,521,288** |                         1.92                          |             |                         **0**                         |

@@

```{r}
#| out-height: 600px
regional_charts <- tar_read(regional_charts)
regional_charts[1]
```

@@

```{r}
#| out-height: 600px
regional_charts[2]
```

@@

```{r}
#| out-height: 600px
regional_charts[3]
```

@@

```{r}
#| out-height: 600px
regional_charts[4]
```

@@

```{r}
#| out-height: 600px
regional_charts[5]
```

@@

```{r}
#| out-height: 600px
regional_charts[6]
```

@@

```{r}
#| out-height: 600px
regional_charts[7]
```
:::

@@

The methodology used for this metric follows, as far as possible, that detailed by NHS Digital[^7]. The readmissions counted relate to those who had an elective PCI / CABG in the period 1/3/2022 to 28/2/2023.

[^7]: <https://digital.nhs.uk/data-and-information/publications/statistical/ccg-outcomes-indicator-set/specifications/3.2-emergency-readmissions-within-30-days-of-discharge-from-hospital_1_4>

@@

Public Health England (PHE) CHD synthetic prevalence estimates data was previously available from OHID via Fingertips. This data has now been removed from Fingertips and archived, however OHID supplied it via email for this project following a request. This data is for 7564 practices from 2015, of which 6297 are current practices. In order to assign a prevalence for practices with missing data a nearest neighbours methodology was used to impute the prevalence by taking an average of the prevalence of each practice within a 1.5km radius of the practice without data. Further details, including code with sample data, for this methodology are available in a blog post on the Strategy Unit Data Science website[^8].

[^8]: <https://the-strategy-unit.github.io/data_science/blogs/posts/2024-01-17_nearest_neighbour.html>

@@

Since metric data is mainly from the 2022/23 year, the October 2022 list size data was taken from the NHS Digital Patients Registered at a GP practice dataset[^9] as this is the data closest to the mid point in the 22/23 year. The GP to ICB mapping file was also downloaded and joined to the list size to easily create analysis by ICB and region.

[^9]: <https://digital.nhs.uk/data-and-information/publications/statistical/patients-registered-at-a-gp-practice/october-2022>

@@

The 2021 Census Ethnicity data was downloaded from the .gov.uk website.[^10] Summary details of the census ethnicity data is also available and this includes descriptions of the standardised list of ethnic groups and the differences from the 2011 ethnic groups.[^11]

[^10]: <https://www.ethnicity-facts-figures.service.gov.uk/uk-population-by-ethnicity/demographics/age-groups/latest/#download-the-data>

[^11]: <https://www.ethnicity-facts-figures.service.gov.uk/uk-population-by-ethnicity/demographics/age-groups/latest/>

@@

Further details about item editing and imputation processes for Census 2021, England and Wales are available on the ONS website. This details techniques used by ONS with the aim of arriving at a fully populated clean database. These techniques include manual imputation and nearest neighbour donor imputation. For the England and Wales Census 2021 the imputation rate for ethnic group was 1.3%.[^12]

[^12]: <https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/methodologies/itemeditingandimputationprocessforcensus2021englandandwales>

@@

The K-medoids clustering was performed using the {cluster} package[^13] in R using partitioning (clustering) of the data into k clusters "around medoids" (PAM), which is a more robust version of K-means. The algorithm is based on the search for k representative objects or medoids among the observations of the dataset. After finding a set of k medoids, k clusters are constructed by assigning each observation to the nearest medoid. The goal is to find k representative objects which minimise the sum of the dissimilarities of the observations to their closest representative object.[^14]

[^13]: Maechler M, Rousseeuw P, Struyf A, Hubert M, Hornik K (2023). *cluster: Cluster Analysis Basics and Extensions*. R package version 2.1.6 — For new features, see the 'NEWS' and the 'Changelog' file in the package source), [https://CRAN.R-project.org/package=cluster](https://cran.r-project.org/package=cluster).

[^14]: <https://cran.r-project.org/web/packages/cluster/cluster.pdf>

@@

A report by Pearcy and Keppel (2002) presents the Index of Disparity as a summary measure of disparity across population groups and advocates its use for groups defined in terms of categories such as ethnicity.[^15]

[^15]: <https://journals.sagepub.com/doi/epdf/10.1093/phr/117.3.273>

@@

> "Practitioners need to be aware that statistical significance differs from practical importance in that statistical significance is highly dependent upon sample size. For a large sample, a statistically significant result is likely to be obtained even when the actual magnitude of an effect is small and of little or no practical importance. On the other hand, for a small sample, it is quite likely that insufficient evidence of a statistically significant result will be obtained – even when there is, indeed, an effect of practical importance." (Hahn, G. J., Doganaksoy, N., Meeker, W. Q., 2019)[^16]

[^16]: <https://academic.oup.com/jrssig/article/16/4/20/7038025>

@@

An explanation of bootstrap theory and confidence intervals is available in Practical Statistics for Data Scientists (Bruce, P. and Bruce, A., 2017)[^17]

[^17]: <https://www.oreilly.com/library/view/practical-statistics-for/9781491952955/>

@@

| Cluster       | CHD Prevalence (%) |  List size | CHD Prevalence | CHD Deaths (% of list size) | Activity rate (CHD Deaths/Prevalence) |
|------------|:----------:|-----------:|-----------:|:----------:|:-------------:|
| Least diverse |        8.7%        | 14,000,000 |      1,218,000 |        6000 (0.043%)        |                0.00492                |
| Most diverse  |       10.6%        |  4,000,000 |        424,000 |        1714 (0.043%)        |                0.00404                |

@@

| Cluster | Population | ASMR - All cause | ASMR - CHD | Rate - All Cause (All deaths / Prevalence) | Rate - CHD (CHD deaths / Prevalence) |
|:---------:|----------:|----------:|----------:|---------------:|------------:|
|    1    | 16,914,882 |            913.5 |       93.9 |                                     1188.4 |                                123.1 |
|    2    | 15,788,339 |            894.3 |       87.6 |                                     1029.5 |                                100.5 |
|    3    | 16,416,864 |            935.4 |       93.3 |                                      813.7 |                                 80.6 |
|    4    |  7,883,280 |            773.7 |       82.2 |                                      431.4 |                                 45.9 |
|    5    |  4,961,545 |            868.2 |      105.8 |                                      511.1 |                                 62.7 |

@@

In July 2021 ONS produced an experimental analysis of ethnic differences in life-expectancy and cause-specific mortality rates[^18]. The analysis found that White and Mixed ethnic groups had lower life expectancy than all other ethnic groups, while the Black African ethnic group had statistically significant higher life expectancy than most groups. However, the analysis also found that ischaemic heart disease mortality was highest in the Bangladeshi, Pakistani and Indian ethnic groups and lowest among Black ethnic groups. The analysis revealed complex patterns in life expectancy and mortality by ethnic group. Potential reasons are offered for the higher life expectancy found in Black African and Asian Other ethnic groups including that they contain a greater proportion of more recent migrants than other ethnic groups and it also cites previous research showing that people who migrate tend to be healthier than others. The ONS report also suggests that Asian and Black ethnic groups engage less in harmful health-related behaviours. These findings show some similarities to research in the USA into what has been termed the Hispanic paradox, where most Hispanic groups are characterised by low economic status and a high prevalence of cardiovascular disease risk factors, but better than expected health and mortality outcomes[^19]. Possible explanations for the Hispanic paradox include the healthy migrant hypothesis (those who migrate tend to be younger and healthier), the presence of nuclear families and high levels of social support, and dietary factors.

[^18]: <https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/lifeexpectancies/articles/ethnicdifferencesinlifeexpectancyandmortalityfromselectedcausesinenglandandwales/2011to2014>

[^19]: <https://www.sciencedirect.com/science/article/abs/pii/S0033062014001339?via%3Dihub>

@@

All the code used in this project is available on the Strategy Unit GitHub[^20].

[^20]: <https://github.com/The-Strategy-Unit/bhf_chd-ethnicity>
